Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-12-11 paper.tex
Tone down banter
Reviewer A:
  - p1, l24 “zone à régime restrictive”: comme vous écrivez en anglais,
  il vaut peut-être mieux éviter les blagues qui ne seront pas
  comprises à l'international. C'est parfaitement à propos dans un
  contexte franco-français, mais je ne pense pas que ce soit une
  supposition que vous devriez faire.

-> removed

Reviewer B:

  I find the tone is trying too hard to be unnecessarily goofy / saturday-morning-cartoony
  in places (lines 18-52 for instance), [culminating in Footnote 3 which made me physically cringe.]

-> rewritten

Anyone who has ever read a cryptography paper knows this for a fact:
cryptographers are at the top of the academic thought chain. In the
archetypal symmetric cryptography paper, there is a ``Section 2''
scene in which cryptographers are bravely waving their most potent
finite field. In a latter ``Section 4'' scene, the very same
cryptographers are seen extolling the virtues of some nifty
\textsc{Arm} Cortex assembly instruction, which takes center stage in
their hand-crafted, manually optimized implementation. No wonder we
jealously stash them away in ``zone à régime restrictive''! The rumor
goes that, behind these padlocked doors, they have developed a set of
domain-specific techniques to deliver high-throughput software
implementation of symmetric ciphers. This has been slowly gnawing at
compiler writers, whose job security is now
threatened~\citep{bernstein:death-compiler,venkatesh:compiler-security}.
Luckily for us, every now and then, someone inadvertently let the
genies out of their Klein bottle and we get to read about the subtle
art of ``bitslicing''~\citep{biham:DES} and
``fixslicing''~\citep{adominicai:fixslicing-gift,
  adomnicai:fixslicing-AES-like}.

The present work aims at casting these two notions in the language of
functional programming. We shall see that bitslicing corresponds to a
run-off-the-mill \emph{data} representation change (in effect, a
matrix transposition). But this is only half of the work: our ability
to systematically transform the \emph{code} operating on such
representation rests on suitably polymorphic definitions:
parametricity at its best! Fixslicing, on the other hand, will be
rationalized through equational reasoning over purely functional
terms. Identifying commuting expressions is of key importance there:
programming with algebraic structures (rather than untamed arrays of
bits) will be instrumental!

In the long run, our ambition is to offer cryptographers the
conceptual apparatus to better illuminate their ``Section 4'' scene.
Each and every symmetric cryptosystem should not have to exhibit a
bitsliced or fixsliced implementation: by just \emph{specifying} the
cryptosystem in the right framework (abstract yet suitably
operational), one should get the guarantee that the design can be
bitsliced and fixsliced. In fact, one would boldly claim that a
compiler could automatically apply these transformations and produce
optimized code, starting from an high-level description of the cipher.
This is already the case for bitslicing~\citep{mercadier:PhD}, adding support
for fixslicing is next on the list.

@@

As intimidating as cryptographers are, the essence of modern symmetric
cryptographic primitives can in fact be distilled in a few paragraph. A
primitive can be understood as a purely functional program. It takes
two inputs: the ``cleartext'' and a list of ``round subkeys''
(which are derived from a single key through a non-performance-critical process called ``key schedule'' that we shall ignore here).
The output is the ``ciphertext''. The cleartext, the ciphertext and
each individual round subkeys are processed as ``blocks'' of binary
data. The block size is of the order of a hundred bits (typically, 64, 128
or 256 bits).

>>>>

Symmetric cryptographic primitives offer a fascinating blend of
advanced mathematics ---upon which rests their theoretical security
guarantees--- as well as low-level programming ---upon which rests
their efficiency but also the absence of side-channel
attacks~\citep{leroy:side-channel} and resistance to
fault-injection~\citep{heydemann:fault-attack}.
%
In the archetypal symmetric cryptography paper, there is a ``Section
2'' in which the authors justify the theoretical properties of their
cipher by appealing to a carefully chosen finite field. In a latter
``Section 4'', the authors describe their manually optimized
implementation, usually written directly in
assembly~\citep{almeida:jasmin}, and report on their performance in
terms of CPU cycles per encrypted byte.
%
Furthermore, cryptographers have developed a set of domain-specific
techniques to deliver high-throughput software implementation of
symmetric ciphers, such as ``bitslicing''~\citep{biham:DES} and
``fixslicing''~\citep{adominicai:fixslicing-gift,
  adomnicai:fixslicing-AES-like}. So far, these techniques have
remained at odds with off-the-shelf compiler
technology~\citep{bernstein:death-compiler,venkatesh:compiler-security}.

The present work aims at casting symmetric cryptographic primitives in
the language of functional programming. In the process, we shall give
a formal treatment of bitslicing and fixslicing as systematic
transformations of functional programs. We shall see that bitslicing
corresponds to a run-off-the-mill \emph{data} representation change
(in effect, a matrix transposition). But this is only half of the
work: our ability to systematically transform the \emph{code}
operating on such representation rests on suitably polymorphic
definitions: parametricity at its best. Fixslicing, on the other hand,
will be rationalized through equational reasoning over purely
functional terms. Identifying commuting expressions is of key
importance there: programming with algebraic structures (rather than
untamed arrays of bits) will be instrumental.

In the long run, our ambition is to offer cryptographers the
conceptual apparatus to streamline their ``Section 4''.
Each and every symmetric cryptosystem should not have to exhibit a
bitsliced or fixsliced implementation: by just \emph{specifying} the
cryptosystem in the right framework (abstract yet suitably
operational), one should get the guarantee that the design can be
bitsliced and fixsliced. In fact, one would boldly claim that a
compiler could automatically apply these transformations and produce
optimized code, starting from an high-level description of the cipher.
This is already the case for bitslicing~\citep{mercadier:PhD}, adding support
for fixslicing is next on the list.

@@

For our present purposes, the essence of modern symmetric
cryptographic primitives can be distilled in a few paragraph. A
primitive can be understood as a purely functional program. It takes
two inputs: the ``cleartext'' and a list of ``round subkeys'' (which
are derived from a single key through a non-performance-critical
process called ``key schedule'' that we shall ignore here). The output
is the ``ciphertext''. The cleartext, the ciphertext and each
individual round subkeys are processed as ``blocks'' of binary data.
The block size is of the order of a hundred bits (typically, 64, 128
or 256 bits).
